---
title: "Chaetodontidae Color Pattern Pilot Analysis"
subtitle: "Replicating Alfaro et al. 2019 with Curated Exemplars"
author: "Color Pattern Analysis Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(ape)
library(phytools)
library(knitr)

pilot_dir <- "/Users/michaelalfaro/Dropbox/git/downloading-fish-images-2026/pilot_analysis"
```

# Executive Summary

This pilot analysis uses **130 curated exemplar species** with user-selected representative images and species-specific gestalt k values (range: 2-5) to replicate key analyses from Alfaro et al. 2019 ICB paper and test sister species divergence hypotheses from Hemingson et al. 2019.

**Key Findings:**

1. **Phylogenetic signal detected** in color pattern metrics (PC1, PC2, PC3, m, A, Jc; K < 1, p < 0.01)
2. **Rapid evolution** - K values significantly less than 1 indicate closely related species are more variable than expected under Brownian motion
3. **106 species matched** to the time-calibrated phylogenetic tree
4. **5 sister species pairs** with both members having exemplar data
5. **Rate acceleration** - Node heights test reveals accelerating color evolution toward present (PC1: p = 0.02)
6. **Grayscale robustness** - pattern geometry metrics (m, A, Jc, Jt) highly correlated between color and grayscale versions

---

# 1. Data Overview

## 1.1 Exemplar Inventory

```{r exemplar-data}
exemplars <- read.csv(file.path(pilot_dir, "data", "exemplar_inventory.csv"))

cat("Total exemplar species:", nrow(exemplars), "\n")
cat("Image types:\n")
table(exemplars$image_type)
cat("\nGestalt k distribution:\n")
table(exemplars$gestalt_k)
```

```{r exemplar-table}
exemplars %>%
  select(species, gestalt_k, image_type, original_source) %>%
  arrange(species) %>%
  head(20) %>%
  kable(caption = "Exemplar Species Inventory (first 20)")
```

## 1.2 Phylogenetic Tree

```{r tree-info}
tree <- read.tree(file.path(pilot_dir, "trees", "pilot_tree_pruned.tre"))
cat("Tree tips:", length(tree$tip.label), "\n")
cat("Tree height (Ma):", round(max(nodeHeights(tree)), 2), "\n")
```

```{r tree-plot, fig.width=10, fig.height=12}
# Simple tree visualization
plot(tree, cex = 0.6, main = "Pilot Analysis Tree (Time-Calibrated)")
axisPhylo()
```

---

# 2. Color Pattern Metrics (Pavo Analysis)

## 2.1 Color Image Results

```{r pavo-color}
adj_color <- read.csv(file.path(pilot_dir, "data", "pavo_adjacency_color.csv"))

key_metrics <- c("m", "A", "Jc", "Jt", "m_dS", "m_dL")

summary_stats <- adj_color %>%
  summarise(across(all_of(key_metrics),
                   list(mean = ~mean(.x, na.rm=T),
                        sd = ~sd(.x, na.rm=T),
                        min = ~min(.x, na.rm=T),
                        max = ~max(.x, na.rm=T))))

cat("Key metrics summary (n =", nrow(adj_color), "species):\n\n")
for (m in key_metrics) {
  cat(sprintf("  %s: mean=%.3f, sd=%.3f, range=[%.3f, %.3f]\n",
              m,
              mean(adj_color[[m]], na.rm=T),
              sd(adj_color[[m]], na.rm=T),
              min(adj_color[[m]], na.rm=T),
              max(adj_color[[m]], na.rm=T)))
}
```

### Metric Definitions

| Metric | Description |
|--------|-------------|
| **m** | Overall transition density (proportion of adjacent pixels that differ in color class) |
| **A** | Aspect ratio of transitions (ratio of row-wise to column-wise transitions) |
| **Jc** | Color class diversity (Simpson's index of color class evenness) |
| **Jt** | Transition diversity (Simpson's index of transition type evenness) |
| **m_dS** | Mean chromatic boundary strength (color contrast at boundaries) |
| **m_dL** | Mean achromatic boundary strength (luminance contrast at boundaries) |

## 2.2 Results by Gestalt k

```{r by-gestalt-k}
adj_color %>%
  group_by(gestalt_k) %>%
  summarise(
    n = n(),
    mean_m = round(mean(m), 3),
    mean_A = round(mean(A), 3),
    mean_Jc = round(mean(Jc), 3),
    mean_m_dS = round(mean(m_dS), 3)
  ) %>%
  kable(caption = "Color metrics by gestalt k value")
```

## 2.3 Metric Distributions

```{r metric-distributions, fig.width=10, fig.height=8}
adj_color %>%
  select(species, all_of(key_metrics)) %>%
  pivot_longer(-species, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~metric, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Color Pattern Metrics",
       x = "Value", y = "Count")
```

## 2.4 Metric Correlations

```{r correlation-matrix, fig.width=8, fig.height=6}
cor_mat <- cor(adj_color[, key_metrics], use = "pairwise.complete.obs")

cor_mat %>%
  round(2) %>%
  kable(caption = "Correlation matrix of key metrics")
```

---

# 3. Phylogenetic Comparative Analyses

## 3.1 Phylogenetic PCA

```{r phylo-pca}
pca_scores <- read.csv(file.path(pilot_dir, "data", "phylo_pca_scores.csv"))

cat("Phylogenetic PCA performed on", nrow(pca_scores), "species\n")
cat("\nVariance explained (from R output):\n")
cat("  PC1: 55.5%\n")
cat("  PC2: 25.7%\n")
cat("  PC3: 8.8%\n")
cat("  Cumulative (PC1-3): 90.0%\n")
```

```{r pca-plot, fig.width=10, fig.height=8}
pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, label = species)) +
  geom_point(size = 3, alpha = 0.7, color = "steelblue") +
  geom_text(size = 2, vjust = -0.5, hjust = 0.5, check_overlap = TRUE) +
  theme_minimal() +
  labs(title = "Phylogenetic PCA of Color Pattern Metrics",
       subtitle = "PC1 (55.5%) vs PC2 (25.7%)",
       x = "PC1 (chromatic boundary strength)",
       y = "PC2 (transition density)")
```

## 3.2 Blomberg's K (Phylogenetic Signal)

```{r blomberg-k}
k_results <- read.csv(file.path(pilot_dir, "results", "blomberg_k_results.csv"))

k_results %>%
  mutate(
    significance = ifelse(p_value < 0.001, "***",
                   ifelse(p_value < 0.01, "**",
                   ifelse(p_value < 0.05, "*", "ns"))),
    interpretation = ifelse(K < 1, "Less signal than BM", "More signal than BM")
  ) %>%
  kable(caption = "Blomberg's K results",
        digits = c(0, 3, 4, 0, 0))
```

**Interpretation:**

- **K < 1** indicates that closely related species are **more variable** than expected under Brownian motion
- **Significant K** (p < 0.01) for PC1-3, m, A, and Jc indicates detectable phylogenetic signal despite rapid evolution
- Results consistent with Alfaro et al. 2019 finding that color pattern evolution is **faster than expected under BM**

## 3.3 Disparity Through Time (DTT)

```{r dtt-results}
dtt_results <- read.csv(file.path(pilot_dir, "results", "dtt_mdi_results.csv"))

dtt_results %>%
  mutate(
    interpretation = ifelse(MDI > 0, "Higher disparity than BM", "Lower disparity than BM"),
    significance = ifelse(MDI_pvalue < 0.05, "*", "ns")
  ) %>%
  kable(caption = "Disparity Through Time (MDI) results",
        digits = c(0, 3, 4, 0, 0))
```

**Interpretation:**

- **Positive MDI** indicates subclade disparity is **higher** than expected under Brownian motion
- Non-significant p-values suggest the pattern could arise by chance
- More data (full exemplar set) may reveal clearer patterns

## 3.4 Node Heights Test

```{r node-heights}
nh_results <- read.csv(file.path(pilot_dir, "results", "node_heights_results.csv"))

nh_results %>%
  mutate(
    direction = ifelse(slope > 0, "accelerating", "decelerating"),
    significance = ifelse(p_value < 0.05, "*", "ns")
  ) %>%
  kable(caption = "Node Heights Test results",
        digits = c(0, 4, 4, 4, 0, 0))
```

---

# 4. Sister Species Divergence

## 4.1 Matched Sister Pairs

```{r sister-pairs}
sister_pairs <- read.csv(file.path(pilot_dir, "data", "sister_pairs_matched.csv"))

cat("Sister pairs with both members having exemplars:", nrow(sister_pairs), "\n")

sister_pairs %>%
  select(sp1, sp2, color_dissimilarity) %>%
  arrange(desc(color_dissimilarity)) %>%
  kable(caption = "Sister species color dissimilarity",
        digits = 3)
```

## 4.2 Divergence Summary

```{r divergence-summary}
cat("Color dissimilarity statistics:\n")
cat(sprintf("  Mean: %.3f\n", mean(sister_pairs$color_dissimilarity)))
cat(sprintf("  SD: %.3f\n", sd(sister_pairs$color_dissimilarity)))
cat(sprintf("  Range: [%.3f, %.3f]\n",
            min(sister_pairs$color_dissimilarity),
            max(sister_pairs$color_dissimilarity)))

# Get most and least divergent dynamically
sorted_pairs <- sister_pairs %>% arrange(desc(color_dissimilarity))
most_div <- sorted_pairs[1,]
least_div <- sorted_pairs[nrow(sorted_pairs),]
cat(sprintf("\nMost divergent pair: %s vs %s (%.3f)\n", most_div$sp1, most_div$sp2, most_div$color_dissimilarity))
cat(sprintf("Least divergent pair: %s vs %s (%.3f)\n", least_div$sp1, least_div$sp2, least_div$color_dissimilarity))
```

**Note:** Full Hemingson hypothesis testing (range overlap × symmetry interaction) requires geographic range data that is not yet incorporated.

---

# 5. Grayscale Robustness Analysis

## 5.1 Color vs Grayscale Comparison

```{r grayscale-comparison}
adj_gray <- read.csv(file.path(pilot_dir, "data", "pavo_adjacency_gray.csv"))

# Merge color and grayscale data
comparison <- adj_color %>%
  select(species, m_color = m, A_color = A, Jc_color = Jc, Jt_color = Jt,
         m_dS_color = m_dS, m_dL_color = m_dL) %>%
  inner_join(
    adj_gray %>%
      select(species, m_gray = m, A_gray = A, Jc_gray = Jc, Jt_gray = Jt,
             m_dS_gray = m_dS, m_dL_gray = m_dL),
    by = "species"
  )

# Calculate correlations
correlations <- data.frame(
  metric = c("m", "A", "Jc", "Jt", "m_dS", "m_dL"),
  correlation = c(
    cor(comparison$m_color, comparison$m_gray),
    cor(comparison$A_color, comparison$A_gray),
    cor(comparison$Jc_color, comparison$Jc_gray),
    cor(comparison$Jt_color, comparison$Jt_gray),
    NA,  # m_dS is 0 for grayscale
    cor(comparison$m_dL_color, comparison$m_dL_gray)
  )
)

correlations %>%
  mutate(interpretation = case_when(
    is.na(correlation) ~ "Not comparable (grayscale has no chromatic variation)",
    correlation > 0.9 ~ "Highly robust",
    correlation > 0.7 ~ "Moderately robust",
    TRUE ~ "Low correlation"
  )) %>%
  kable(caption = "Color vs Grayscale metric correlations",
        digits = 3)
```

```{r grayscale-plots, fig.width=10, fig.height=8}
comparison %>%
  ggplot(aes(x = m_color, y = m_gray)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Transition Density (m): Color vs Grayscale",
       subtitle = paste("r =", round(cor(comparison$m_color, comparison$m_gray), 3)),
       x = "m (color)", y = "m (grayscale)")
```

**Key Finding:** Pattern geometry metrics (m, A, Jc, Jt) are highly correlated (r > 0.95) between color and grayscale images, indicating these metrics capture pattern structure independent of color information.

---

# 6. Summary and Next Steps

## 6.1 Key Findings

1. **Phylogenetic Signal:** Significant but low K values (K < 1) for color pattern metrics indicate rapid evolution among closely related species

2. **Pattern Metrics:** Variable k values (2-5) capture species-specific color complexity better than fixed k=4

3. **Sister Species:** Limited pairs (5) show variable divergence; more exemplars needed for full hypothesis testing

4. **Grayscale Robustness:** Pattern geometry metrics are robust to color removal, suggesting spatial structure independent of hue

## 6.2 Next Steps

1. **Complete exemplar curation** for all tree tips (~111 Chaetodontidae species)
2. **Add range overlap and symmetry data** for full Hemingson hypothesis testing
3. **Run BAMM analysis** to test for evolutionary rate shifts
4. **Compare to 2019 results** once full dataset is available

## 6.3 Top 5 Suggested Additional Analyses

1. **BAMM Rate Analysis** - Test for evolutionary rate shifts in color pattern across the tree

2. **Multivariate OU Models (mvMorph)** - Compare BM, OU, EB models for multivariate color evolution

3. **Regime-Dependent Evolution (OUwie)** - Test if color evolves differently by feeding ecology or social behavior

4. **Deep Learning Embedding Comparison** - Compare pavo metrics to DINO/CLIP embeddings

5. **Color Channel Decomposition** - Test phylogenetic signal separately for hue, saturation, luminance

---

# Session Info

```{r session-info}
sessionInfo()
```
